### 2025-07-19 paper/quarto_aaai/paper.tex
adjusted Algorithm 1 and ensured consistet capitalization for CT, AR, CE etc.
trying to address comments on figure 2

The remainder of this manuscript is structured as follows.
Section~\ref{sec-lit} presents related work, focusing on the links to
contrastive and robust learning. Then follow our two principal
contributions. In Section~\ref{sec-method}, we introduce our
methodological framework and show theoretically that it can be employed
to enforce global actionability constraints. In our experiments
(Section~\ref{sec-experiments}), we find that thanks to counterfactual
training, (1) the implausibility of CEs decreases by up to 90\%; (2) the
cost of reaching valid counterfactuals with protected features decreases
by 19\% on average; and (3) the adversarial robustness improves across
the board. Finally, we discuss open challenges in
Section~\ref{sec-discussion} and conclude in
Section~\ref{sec-conclusion}.

@@

\begin{algorithm}[h]
  \caption{Counterfactual Training}
    \label{alg-experiment}
    \begin{algorithmic}[1]
    \REQUIRE Training dataset $\mathcal{D}$, initial model $\mathbf{M}_{\theta_0}$
    \WHILE{not converged}
        \STATE Sample $\mathbf{x}$ and $\mathbf{y}$ from dataset $\mathcal{D}$.
        \STATE Sample $\mathbf{x}^{\prime}_0$, $\mathbf{y}^+$ and $\mathbf{x}^+$.
        \FOR{$t = 1$ to $T$}
            \STATE Backpropagate $\nabla_{\mathbf{x}^\prime}$ through Equation \ref{eq-general}. Store $\mathbf{x}_t^\prime$.
        \ENDFOR
        \STATE Backpropagate $\nabla_{\theta}$ through Equation \ref{eq-obj}.
    \ENDWHILE
    \RETURN $\mathbf{M}_\theta$
    \end{algorithmic}
\end{algorithm}

@@

The departure from SGLD allows us to tap into the vast repertoire of
explainers that have been proposed in the literature to meet different
desiderata. For example, many methods support domain and mutability
constraints. In principle, any existing approach for generating CEs is
viable, so long as it does not violate the faithfulness condition. Like
JEMs \citep{murphy2022probabilistic}, Counterfactual Training can be
considered a form of contrastive representation learning.

@@

We run experiments with three gradient-based generators: \emph{Generic}
of \citet{wachter2017counterfactual} as a simple baseline approach,
\emph{REVISE} \citep{joshi2019realistic} that aims to generate plausible
counterfactuals using a surrogate Variational Autoencoder (VAE), and
\emph{ECCo}---the generator of \citet{altmeyer2024faithful} without the
conformal prediction component---as a method that directly targets both
faithfulness and plausibility of the counterfactuals.

@@

\caption{\label{tbl-main}Key evaluation metrics for valid counterfactual
along with bootstrapped standard errors for all datasets.
\textbf{Plausibility} (columns 1-2): percentage reduction in
implausibility for \(\text{IP}\) and \(\text{IP}^*\), respectively;
\textbf{Cost} / \textbf{Actionability} (column 3): percentage reduction
in costs when selected features are protected. Outcomes are aggregated
across bootstrap samples and varying degrees of the energy penalty
\(\lambda_{\text{egy}}\) used for \emph{ECCo} at test time. Asterisks
(\(^*\)) indicate that the bootstrapped 99\%-confidence interval of
differences in mean outcomes does \emph{not} include zero.}

@@

\emph{CT generally produces substantial and statistically significant
improvements in plausibility.}\\
Average reductions in \(\text{IP}\) range from around 7\% for
\emph{MNIST} to almost 60\% for \emph{Circ}. For the real-world tabular
datasets they are around 12\% for \emph{CH} and \emph{Cred} and almost
25\% for \emph{GMSC}; for \emph{Adult} and \emph{OL} we find no
significant impact of CT on \(\text{IP}\). Reductions in \(\text{IP}^*\)
are even more substantial and generally statistically significant,
although the average degree of uncertainty is higher than for
\(\text{IP}\): reductions range from around 20\% (\emph{Moons}) to
almost 90\% (\emph{Circ}). The only negative findings for \emph{OL} and
\emph{MNIST} are not statistically significant and, for MNIST, do not
align with qualitative findings, which are much more plausible for CT
(Figure~\ref{fig-mnist}).

@@

\caption{\label{fig-mnist}Sample explanations for \emph{MNIST} for BL
(top) and CT (bottom). First column is a random factual 0 (blue).
Columns 2 to 5 are corresponding \emph{ECCo} counterfactuals in target
classes 1 to 4. Columns 6 to 10 show integrated gradients averaged over
all test images in classes 5 to 9.}

@@

For synthetic datasets, we always protect the first feature; for all
real-world tabular datasets we could identify and protect an \emph{age}
variable; for \emph{MNIST}, we protect the five upper and lower pixel
rows of the full image. Reductions in costs are overwhelmingly positive
and significant, of up to around 60\% for \emph{GMSC}. While the
estimated cost increase for \emph{Adult} and \emph{MNIST}, these effects
are not significant and Figure~\ref{fig-mnist} (columns 6-10)
demonstrates that in the latter case, CT does have the expected effect:
sensitivity to protected features as per integrated gradients is
drastically reduced; details of this experiment are reported in the
supplementary appendix. In the case of \emph{Cred}, average costs
increase, likely because any potential benefits from protecting the
\emph{age} are outweighed by the increase in costs required for greater
plausibility.

>>>>

The remainder of this manuscript is structured as follows.
Section~\ref{sec-lit} presents related work, focusing on the links to
contrastive and robust learning. Then follow our two principal
contributions. In Section~\ref{sec-method}, we introduce our
methodological framework and show theoretically that it can be employed
to enforce global actionability constraints. In our experiments
(Section~\ref{sec-experiments}), we find that thanks to counterfactual
training, (1) the implausibility of CEs decreases by up to 90\%; (2) the
cost of reaching valid counterfactuals with protected features decreases
by 19\% on average; and (3) models' adversarial robustness improves
across the board. Finally, we discuss open challenges in
Section~\ref{sec-discussion} and conclude in
Section~\ref{sec-conclusion}.

@@

\begin{algorithm}[h]
  \caption{Counterfactual Training}
    \label{alg-experiment}
    \begin{algorithmic}[1]
    \REQUIRE Training dataset $\mathcal{D}$, initialize model $\mathbf{M}_{\theta}$
    \WHILE{not converged}
        \STATE Sample $\mathbf{x}$ and $\mathbf{y}$ from dataset $\mathcal{D}$.
        \STATE Sample $\mathbf{x}^{\prime}_0$, $\mathbf{y}^+$ and $\mathbf{x}^+$.
        \FOR{$t = 1$ to $T$}
            \STATE Backpropagate $\nabla_{\mathbf{x}^\prime}$ through Equation \ref{eq-general}. Store $\mathbf{x}_t^\prime$.
        \ENDFOR
        \STATE Backpropagate $\nabla_{\theta}$ through Equation \ref{eq-obj}.
    \ENDWHILE
    \RETURN $\mathbf{M}_\theta$
    \end{algorithmic}
\end{algorithm}

@@

The departure from SGLD allows us to tap into the vast repertoire of
explainers that have been proposed in the literature to meet different
desiderata. For example, many methods support domain and mutability
constraints. In principle, any existing approach for generating CEs is
viable, so long as it does not violate the faithfulness condition. Like
JEMs \citep{murphy2022probabilistic}, counterfactual training can be
considered a form of contrastive representation learning.

@@

We run experiments with three gradient-based generators: \emph{Generic}
of \citet{wachter2017counterfactual} as a simple baseline approach,
\emph{REVISE} \citep{joshi2019realistic} that aims to generate plausible
counterfactuals using a surrogate Variational Autoencoder (VAE), and
\emph{ECCCo} \citep{altmeyer2024faithful}, which targets faithfulness.

@@

\caption{\label{tbl-main}Key evaluation metrics for valid counterfactual
along with bootstrapped standard errors for all datasets.
\textbf{Plausibility} (columns 1-2): percentage reduction in
implausibility for \(\text{IP}\) and \(\text{IP}^*\), respectively;
\textbf{Cost} / \textbf{Actionability} (column 3): percentage reduction
in costs when selected features are protected. Outcomes are aggregated
across bootstrap samples and varying degrees of the energy penalty
\(\lambda_{\text{egy}}\) used for \emph{ECCCo} at test time. Asterisks
(\(^*\)) indicate that the bootstrapped 99\%-confidence interval of
differences in mean outcomes does \emph{not} include zero.}

@@

\emph{CT generally produces substantial and statistically significant
improvements in plausibility.}\\
Average reductions in \(\text{IP}\) range from around 7\% for
\emph{MNIST} to almost 60\% for \emph{Circ}. For the real-world tabular
datasets they are around 12\% for \emph{CH} and \emph{Cred} and almost
25\% for \emph{GMSC}; for \emph{Adult} and \emph{OL} we find no
significant impact of CT on \(\text{IP}\). Reductions in \(\text{IP}^*\)
are even more substantial and generally statistically significant,
although the average degree of uncertainty is higher than for
\(\text{IP}\): reductions range from around 20\% (\emph{Moons}) to
almost 90\% (\emph{Circ}). The only negative findings are for OL and
MNIST, but they are not statistically significant. A qualitative
inspection of the counterfactuals in Figure~\ref{fig-mnist} (columns
2-5) suggests recognizable digits 1-4 for the model trained with CT
(bottom row), unlike the baseline (top row).

@@

\caption{\label{fig-mnist}Visual explanations for \emph{MNIST} for BL
(top) and CT (bottom). \textbf{Plausibility}: col.~1 is a random factual
0 (blue); cols. 2-5 are corresponding \emph{ECCCo} counterfactuals in
target classes 1 to 4. \textbf{Actionability}: cols. 6-10 show
integrated gradients averaged over test images in classes 5 to 9.}

@@

For synthetic datasets, we always protect the first feature; for all
real-world tabular datasets we could identify and protect an \emph{age}
variable; for \emph{MNIST}, we protect the five upper and lower pixel
rows of the full image. Statistically significant reductions in costs
overwhelmingly point in the expected positive direction reaching up to
around 60\% for \emph{GMSC}. Only in the case of \emph{Cred}, average
costs increase, likely because any potential benefits from protecting
the \emph{age} are outweighed by the increase in costs required for
greater plausibility. The findings for \emph{Adult} and \emph{MNIST} are
not significant. A qualitative inspection of the class-conditional
integrated gradients in Figure~\ref{fig-mnist} (columns 6-10) suggests
that CT still has the expected effect: the model (bottom) is insensitive
(blue) to the protected rows of pixels; details of this experiment are
reported in the supplementary appendix.

### 2025-07-19 paper/sections/experiments.qmd
trying to address comments on figure 2

We run experiments with three gradient-based generators: *Generic* of @wachter2017counterfactual as a simple baseline approach, *REVISE* [@joshi2019realistic] that aims to generate plausible counterfactuals using a surrogate Variational Autoencoder (VAE), and *ECCo*---the generator of @altmeyer2024faithful without the conformal prediction component---as a method that directly targets both faithfulness and plausibility of the counterfactuals.

@@

::: {#tbl-main}
```{=latex}
\begin{table}[h]
\small
\centering
\input{tables/main.tex}
\end{table}
```
Key evaluation metrics for valid counterfactual along with bootstrapped standard errors for all datasets. **Plausibility** (columns 1-2): percentage reduction in implausibility for $\text{IP}$ and $\text{IP}^*$, respectively; **Cost** / **Actionability** (column 3): percentage reduction in costs when selected features are protected. Outcomes are aggregated across bootstrap samples and varying degrees of the energy penalty $\lambda_{\text{egy}}$ used for *ECCo* at test time. Asterisks ($^*$) indicate that the bootstrapped 99%-confidence interval of differences in mean outcomes does *not* include zero.
:::

### Plausibility (RQ1). {#sec-plaus}
*CT generally produces substantial and statistically significant improvements in plausibility.*  
Average reductions in $\text{IP}$ range from around 7% for *MNIST* to almost 60% for *Circ*. For the real-world tabular datasets they are around 12% for *CH* and *Cred* and almost 25% for *GMSC*; for *Adult* and *OL* we find no significant impact of CT on $\text{IP}$. Reductions in $\text{IP}^*$ are even more substantial and generally statistically significant, although the average degree of uncertainty is higher than for $\text{IP}$: reductions range from around 20% (*Moons*) to almost 90% (*Circ*). The only negative findings for *OL* and *MNIST* are not statistically significant and, for MNIST, do not align with qualitative findings, which are much more plausible for CT (@fig-mnist).

![Sample explanations for *MNIST* for BL (top) and CT (bottom). First column is a random factual 0 (blue). Columns 2 to 5 are corresponding *ECCo* counterfactuals in target classes 1 to 4. Columns 6 to 10 show integrated gradients averaged over all test images in classes 5 to 9.](/paper/figures/mnist_body.png){#fig-mnist}

@@

For synthetic datasets, we always protect the first feature; for all real-world tabular datasets we could identify and protect an *age* variable; for *MNIST*, we protect the five upper and lower pixel rows of the full image. Reductions in costs are overwhelmingly positive and significant, of up to around 60% for *GMSC*. While the estimated cost increase for *Adult* and *MNIST*, these effects are not significant and @fig-mnist (columns 6-10) demonstrates that in the latter case, CT does have the expected effect: sensitivity to protected features as per integrated gradients is drastically reduced; details of this experiment are reported in the supplementary appendix. In the case of *Cred*, average costs increase, likely because any potential benefits from protecting the *age* are outweighed by the increase in costs required for greater plausibility.

>>>>

We run experiments with three gradient-based generators: *Generic* of @wachter2017counterfactual as a simple baseline approach, *REVISE* [@joshi2019realistic] that aims to generate plausible counterfactuals using a surrogate Variational Autoencoder (VAE), and *ECCCo* [@altmeyer2024faithful], which targets faithfulness.

@@

::: {#tbl-main}
```{=latex}
\begin{table}[h]
\small
\centering
\input{tables/main.tex}
\end{table}
```
Key evaluation metrics for valid counterfactual along with bootstrapped standard errors for all datasets. **Plausibility** (columns 1-2): percentage reduction in implausibility for $\text{IP}$ and $\text{IP}^*$, respectively; **Cost** / **Actionability** (column 3): percentage reduction in costs when selected features are protected. Outcomes are aggregated across bootstrap samples and varying degrees of the energy penalty $\lambda_{\text{egy}}$ used for *ECCCo* at test time. Asterisks ($^*$) indicate that the bootstrapped 99%-confidence interval of differences in mean outcomes does *not* include zero.
:::

### Plausibility (RQ1). {#sec-plaus}
*CT generally produces substantial and statistically significant improvements in plausibility.*  
Average reductions in $\text{IP}$ range from around 7% for *MNIST* to almost 60% for *Circ*. For the real-world tabular datasets they are around 12% for *CH* and *Cred* and almost 25% for *GMSC*; for *Adult* and *OL* we find no significant impact of CT on $\text{IP}$. Reductions in $\text{IP}^*$ are even more substantial and generally statistically significant, although the average degree of uncertainty is higher than for $\text{IP}$: reductions range from around 20% (*Moons*) to almost 90% (*Circ*). The only negative findings are for OL and MNIST, but they are not statistically significant. A qualitative inspection of the counterfactuals in @fig-mnist (columns 2-5) suggests recognizable digits 1-4 for the model trained with CT (bottom row), unlike the baseline (top row).

![Visual explanations for *MNIST* for BL (top) and CT (bottom). **Plausibility**: col. 1 is a random factual 0 (blue); cols. 2-5 are corresponding *ECCCo* counterfactuals in target classes 1 to 4. **Actionability**: cols. 6-10 show integrated gradients averaged over test images in classes 5 to 9.](/paper/figures/mnist_body.png){#fig-mnist}

@@

For synthetic datasets, we always protect the first feature; for all real-world tabular datasets we could identify and protect an *age* variable; for *MNIST*, we protect the five upper and lower pixel rows of the full image. Statistically significant reductions in costs overwhelmingly point in the expected positive direction reaching up to around 60% for *GMSC*. Only in the case of *Cred*, average costs increase, likely because any potential benefits from protecting the *age* are outweighed by the increase in costs required for greater plausibility. The findings for *Adult* and *MNIST* are not significant. A qualitative inspection of the class-conditional integrated gradients in @fig-mnist (columns 6-10) suggests that CT still has the expected effect: the model (bottom) is insensitive (blue) to the protected rows of pixels; details of this experiment are reported in the supplementary appendix.

### 2025-07-19 paper/sections/intro.qmd
trying to address comments on figure 2

The remainder of this manuscript is structured as follows. @sec-lit presents related work, focusing on the links to contrastive and robust learning. Then follow our two principal contributions. In @sec-method, we introduce our methodological framework and show theoretically that it can be employed to enforce global actionability constraints. In our experiments (@sec-experiments), we find that thanks to counterfactual training, (1) the implausibility of CEs decreases by up to 90%; (2) the cost of reaching valid counterfactuals with protected features decreases by 19% on average; and (3) the adversarial robustness improves across the board. Finally, we discuss open challenges in @sec-discussion and conclude in @sec-conclusion.

>>>>

The remainder of this manuscript is structured as follows. @sec-lit presents related work, focusing on the links to contrastive and robust learning. Then follow our two principal contributions. In @sec-method, we introduce our methodological framework and show theoretically that it can be employed to enforce global actionability constraints. In our experiments (@sec-experiments), we find that thanks to counterfactual training, (1) the implausibility of CEs decreases by up to 90%; (2) the cost of reaching valid counterfactuals with protected features decreases by 19% on average; and (3) models' adversarial robustness improves across the board. Finally, we discuss open challenges in @sec-discussion and conclude in @sec-conclusion.
